How do changes in ex vary between countries?

### 2022-07-17 draft_working_paper.Rmd
mostly adding Netherlands to comparators

- England & Wales (`GBRTENW`)
- Scotland (`GBR_SCO`)
- UK as a whole (`GBR_UK`) #we haven't included UK but can do
- France (`FRANTP`)
- Spain (`ESP`)
- Italy (`ITA`)
- Germany
   - Total Germany (`DEUTNP`)
   - East Germany (`DEUTE`)
   - West Germany (`DEUTW`)
   - Simulated/Synthetic Germany

@@

# Labels for codes 
country_code_lookup <- 
  tribble(
    ~code, ~country,
    "DEUTNP", "Germany",
    "DEUTE", "East Germany",
    "DEUTW", "West Germany",
    "ESP", "Spain",
    "FRATNP", "France", 
    "ITA", "Italy",
    "GBRTENW", "England & Wales",
    "GBR_SCO", "Scotland",
    "DEUTSYNTH", "Synthetic Germany"
  )

countries_of_interest <- c(
  "GBRTENW",
  "GBR_SCO",
  "GBR_UK",
  "FRATNP",
  "ESP",
  "ITA",
  "DEUTNP",
  "DEUTE", 
  "DEUTW"
)

@@

```{r, echo = FALSE, fig.cap = "Life Expectancy at birth"}
hmd_ex_selected_countries_with_synth %>% 
  left_join(country_code_lookup) %>% 
  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "East Germany", "West Germany", "Germany", "Synthetic Germany", "Spain", "France", "Italy"))) %>% 
  filter(between(year, 1980, 2020)) %>% 
  filter(x == 0) %>% 
  ggplot(aes(x = year, y = ex, group = sex, colour = sex, linetype = sex)) + 
  geom_line() +
  facet_wrap(~country) +
  labs(
    x = "Year",
    y = "Life Expectancy at birth",
    title = "Life expectancies at birth for selected nations",
    subtitle = "1980 to 2020 or latest available year",
    caption = "Source: Human Mortality Database"
  )
```

```{r, echo = FALSE, fig.cap = "Life Expectancy at age 65"}
hmd_ex_selected_countries_with_synth %>% 
  left_join(country_code_lookup) %>% 
  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "East Germany", "West Germany", "Germany", "Synthetic Germany", "Spain", "France", "Italy"))) %>% 
  filter(between(year, 1980, 2020)) %>% 
  filter(x == 65) %>% 
  ggplot(aes(x = year, y = ex, group = sex, colour = sex, linetype = sex)) + 
  geom_line() +
  facet_wrap(~country) +
  labs(
    x = "Year",
    y = "Life Expectancy at age 65",
    title = "Life expectancies at age 65 for selected nations",
    subtitle = "1980 to 2020 or latest available year",
    caption = "Source: Human Mortality Database"
  )
```

@@

```{r, echo = FALSE, warning=FALSE, message = FALSE, fig.height=8, fig.width=8}
change_in_ex_selected_countries %>% 
  filter(x == 0) %>% 
  left_join(country_code_lookup) %>% 
  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "East Germany", "West Germany", "Germany", "Synthetic Germany", "Spain", "France", "Italy"))) %>% 
  filter(between(year, 1980, 2020)) %>% 
  ggplot(aes(x = year, y = delta_ex, colour=sex)) + 
  geom_point(alpha = 0.2) + 
  stat_smooth() + 
  facet_grid(country~sex) + 
  geom_hline(yintercept = 0) + 
  theme(strip.text.y = element_text(angle = 0)) + 
  labs(
    x = "Year",
    y = "Year-on-year change in life expectancy",
    title = "Annual change in life expectancy at birth, selected countries",
    subtitle = "Blue line: nonlinear smoother",
    caption =  "Source: Human Mortality Database. Synthetic Germany based on 20% East/80% West German population weighting"
  )
```

@@

```{r, echo = FALSE, warning=FALSE, message = FALSE, fig.height=8, fig.width=8}
change_in_ex_selected_countries %>% 
  filter(x == 65) %>% 
  left_join(country_code_lookup) %>% 
  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "East Germany", "West Germany", "Germany", "Synthetic Germany", "Spain", "France", "Italy"))) %>% 
  filter(between(year, 1980, 2020)) %>% 
  ggplot(aes(x = year, y = delta_ex, colour=sex)) + 
  geom_point(alpha = 0.2) + 
  stat_smooth() + 
  facet_grid(country~sex) + 
  geom_hline(yintercept = 0) + 
  theme(strip.text.y = element_text(angle = 0)) + 
  labs(
    x = "Year",
    y = "Year-on-year change in life expectancy",
    title = "Annual change in life expectancy at age 65, selected countries",
    subtitle = "Blue line: nonlinear smoother",
    caption =  "Source: Human Mortality Database. Synthetic Germany based on 20% East/80% West German population weighting"
  )
```

@@

```{r, echo = FALSE, warning = FALSE, message = FALSE}
hmd_ex_selected_countries_with_synth %>% 
    filter(year >= 1979) %>% 
    filter(!(code %in% c("DEUTE", "DEUTW", "DEUTNP"))) %>%  # Using only synthetic germany for longer time period
    group_by(code, x, sex) %>% 
    arrange(year) %>% 
    mutate(delta_ex = ex - lag(ex)) %>% 
    nest() %>% 
    mutate(
      rwd_model = map(data, ~lm(delta_ex ~ 1, data = . )),
      ar_model  = map(data, ~lm(delta_ex ~ lag(delta_ex), data = .))
    ) %>% 
    mutate(
      aic_rwd = map_dbl(rwd_model, AIC),
      aic_ar  = map_dbl(ar_model, AIC)
    ) %>% 
    mutate(
      diff_aic = aic_ar - aic_rwd
    )  %>% 
  # filter(diff_aic > 0)
  mutate(
    tidied_ar_model = map(ar_model, broom::tidy)
  ) %>% 
  unnest(tidied_ar_model) %>% 
  filter(term == "(Intercept)") %>% 
  arrange(desc(estimate)) %>%
  left_join(country_code_lookup) %>% 
  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "East Germany", "West Germany", "Germany", "Synthetic Germany", "Spain", "France", "Italy"))) %>% 
  select(code, country, x, sex, estimate, std.error) %>% 
  mutate(display_label = glue::glue("{country}_{x}_{sex}")) %>% 
  ggplot(aes(estimate, fct_reorder(display_label, estimate))) + 
  geom_point(aes(colour = x), show.legend = FALSE) +
  expand_limits( x = 0) + 
  geom_vline(xintercept = 0) + 
  labs(
    x = "Average annual life expectancy change",
    y = "Country-sex combination",
    title = "Average annual improvements by country and sex",
    subtitle = "Range: 1980 to 2020 or nearest available years",
    caption = "Source: Human Mortality Database"
  ) + 
  facet_wrap(~x)

@@

The following combinations of country, starting age, and sex indicate statistically signifanct improvements in model fit if using a model which allows the annual changes to vary by decade

```{r}
all_null_decadal_models %>% 
  filter(anova_pval < 0.05) %>% 
  arrange(desc(anova_pval)) %>% 
  select(code, x, sex, anova_pval)
```
* East Germany, Females, e0
* East Germany, Males, e65

@@

```{r}
all_null_decadal_models %>% 
  filter(anova_pval < 0.05) %>% 
  arrange(desc(anova_pval)) %>% 
  select(code, x, sex, model_alt) %>% 
  mutate(
    model_results = map(model_alt, broom::tidy)
  ) %>% 
  select(-model_alt) %>% 
  unnest(model_results) %>% 
  mutate(
    term  = case_when(
      term == "(Intercept)" ~ "Baseline (1980s)",
      term == "factor(decade)1990" ~ "1990s",
      term == "factor(decade)2000" ~ "2000s",
      term == "factor(decade)2010" ~ "2010s",
      TRUE ~ NA_character_
    )
  ) %>% 
  mutate(
    label = glue::glue("{round(estimate, 3)}{case_when(p.value < 0.01 ~ '***', p.value < 0.05 ~ '**', p.value < 0.10 ~ '*', TRUE ~ '')} ({round(std.error, 3)})")
  ) %>% 
  ungroup() %>% 
  select(
    x, Term = term, label
  ) %>% 
  pivot_wider(values_from = label, names_from = x) %>% 
  kable(format = "html", table.attr = "style=\"color: black;\"",caption = "Coefficients of decadal differences, East Germany, females, by starting age (0 = life expectancy at birth; 65 = life expectancy at age 65 years") %>% 
  kableExtra::kable_styling() %>% 
  footnote(general = "The baseline is the average annual rate of improvement in the 1980s. The other terms indicate increments or decrements from this baseline. Statistical significance of coefficients: * if p < 0.10, ** if p < 0.05, *** if p < 0.01"
  )
```

@@

>>>>

- England & Wales (`GBRTENW`)
- Scotland (`GBR_SCO`)
- UK as a whole (`GBR_UK`) #we haven't included UK but can do
- France (`FRANTP`)
- Spain (`ESP`)
- Italy (`ITA`)
- Germany
   - Total Germany (`DEUTNP`)
   - East Germany (`DEUTE`)
   - West Germany (`DEUTW`)
   - Simulated/Synthetic Germany
- Netherlands (`NLD`) - last minute addition from Danny

@@

# Labels for codes 
country_code_lookup <- 
  tribble(
    ~code, ~country,
    "DEUTNP", "Germany",
    "DEUTE", "East Germany",
    "DEUTW", "West Germany",
    "ESP", "Spain",
    "FRATNP", "France", 
    "ITA", "Italy",
    "GBRTENW", "England & Wales",
    "GBR_SCO", "Scotland",
    "DEUTSYNTH", "Synthetic Germany",
    "NLD", "Netherlands"
  )

countries_of_interest <- c(
  "GBRTENW",
  "GBR_SCO",
  "GBR_UK",
  "FRATNP",
  "ESP",
  "ITA",
  "DEUTNP",
  "DEUTE", 
  "DEUTW",
  "NLD"
)

@@

```{r, echo = FALSE, fig.cap = "Life Expectancy at birth"}
hmd_ex_selected_countries_with_synth %>% 
  left_join(country_code_lookup) %>% 
  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "East Germany", "West Germany", "Germany", "Synthetic Germany", "Spain", "France", "Italy", "Netherlands"))) %>% 
  filter(between(year, 1980, 2020)) %>% 
  filter(x == 0) %>% 
  ggplot(aes(x = year, y = ex, group = sex, colour = sex, linetype = sex)) + 
  geom_line() +
  facet_wrap(~country) +
  labs(
    x = "Year",
    y = "Life Expectancy at birth",
    title = "Life expectancies at birth for selected nations",
    subtitle = "1980 to 2020 or latest available year",
    caption = "Source: Human Mortality Database"
  )
```

```{r, echo = FALSE, fig.cap = "Life Expectancy at age 65"}
hmd_ex_selected_countries_with_synth %>% 
  left_join(country_code_lookup) %>% 
  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "East Germany", "West Germany", "Germany", "Synthetic Germany", "Spain", "France", "Italy", "Netherlands"))) %>% 
  filter(between(year, 1980, 2020)) %>% 
  filter(x == 65) %>% 
  ggplot(aes(x = year, y = ex, group = sex, colour = sex, linetype = sex)) + 
  geom_line() +
  facet_wrap(~country) +
  labs(
    x = "Year",
    y = "Life Expectancy at age 65",
    title = "Life expectancies at age 65 for selected nations",
    subtitle = "1980 to 2020 or latest available year",
    caption = "Source: Human Mortality Database"
  )
```

@@

```{r, echo = FALSE, warning=FALSE, message = FALSE, fig.height=8, fig.width=8}
change_in_ex_selected_countries %>% 
  filter(x == 0) %>% 
  left_join(country_code_lookup) %>% 
  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "East Germany", "West Germany", "Germany", "Synthetic Germany", "Spain", "France", "Italy", "Netherlands"))) %>% 
  filter(between(year, 1980, 2020)) %>% 
  ggplot(aes(x = year, y = delta_ex, colour=sex)) + 
  geom_point(alpha = 0.2) + 
  stat_smooth() + 
  facet_grid(country~sex) + 
  geom_hline(yintercept = 0) + 
  theme(strip.text.y = element_text(angle = 0)) + 
  labs(
    x = "Year",
    y = "Year-on-year change in life expectancy",
    title = "Annual change in life expectancy at birth, selected countries",
    subtitle = "Blue line: nonlinear smoother",
    caption =  "Source: Human Mortality Database. Synthetic Germany based on 20% East/80% West German population weighting"
  )
```

@@

```{r, echo = FALSE, warning=FALSE, message = FALSE, fig.height=8, fig.width=8}
change_in_ex_selected_countries %>% 
  filter(x == 65) %>% 
  left_join(country_code_lookup) %>% 
  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "East Germany", "West Germany", "Germany", "Synthetic Germany", "Spain", "France", "Italy", "Netherlands"))) %>% 
  filter(between(year, 1980, 2020)) %>% 
  ggplot(aes(x = year, y = delta_ex, colour=sex)) + 
  geom_point(alpha = 0.2) + 
  stat_smooth() + 
  facet_grid(country~sex) + 
  geom_hline(yintercept = 0) + 
  theme(strip.text.y = element_text(angle = 0)) + 
  labs(
    x = "Year",
    y = "Year-on-year change in life expectancy",
    title = "Annual change in life expectancy at age 65, selected countries",
    subtitle = "Blue line: nonlinear smoother",
    caption =  "Source: Human Mortality Database. Synthetic Germany based on 20% East/80% West German population weighting"
  )
```

@@

```{r, echo = FALSE, warning = FALSE, message = FALSE}
hmd_ex_selected_countries_with_synth %>% 
    filter(year >= 1979) %>% 
    filter(!(code %in% c("DEUTE", "DEUTW", "DEUTNP"))) %>%  # Using only synthetic germany for longer time period
    group_by(code, x, sex) %>% 
    arrange(year) %>% 
    mutate(delta_ex = ex - lag(ex)) %>% 
    nest() %>% 
    mutate(
      rwd_model = map(data, ~lm(delta_ex ~ 1, data = . )),
      ar_model  = map(data, ~lm(delta_ex ~ lag(delta_ex), data = .))
    ) %>% 
    mutate(
      aic_rwd = map_dbl(rwd_model, AIC),
      aic_ar  = map_dbl(ar_model, AIC)
    ) %>% 
    mutate(
      diff_aic = aic_ar - aic_rwd
    )  %>% 
  # filter(diff_aic > 0)
  mutate(
    tidied_ar_model = map(ar_model, broom::tidy)
  ) %>% 
  unnest(tidied_ar_model) %>% 
  filter(term == "(Intercept)") %>% 
  arrange(desc(estimate)) %>%
  left_join(country_code_lookup) %>% 
  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "East Germany", "West Germany", "Germany", "Synthetic Germany", "Spain", "France", "Italy", "Netherlands"))) %>% 
  select(code, country, x, sex, estimate, std.error) %>% 
  mutate(display_label = glue::glue("{country}_{x}_{sex}")) %>% 
  ggplot(aes(estimate, fct_reorder(display_label, estimate))) + 
  geom_point(aes(colour = x), show.legend = FALSE) +
  expand_limits( x = 0) + 
  geom_vline(xintercept = 0) + 
  labs(
    x = "Average annual life expectancy change",
    y = "Country-sex combination",
    title = "Average annual improvements by country and sex",
    subtitle = "Range: 1980 to 2020 or nearest available years",
    caption = "Source: Human Mortality Database"
  ) + 
  facet_wrap(~x)

@@

The following combinations of country, starting age, and sex indicate statistically significant improvements in model fit if using a model which allows the annual changes to vary by decade

```{r}
all_null_decadal_models %>% 
  filter(anova_pval < 0.05) %>% 
  arrange(desc(anova_pval)) %>% 
  select(code, x, sex, anova_pval)
```
* East Germany, Females, e0
* East Germany, Males, e65
* Netherlands, Males, e65

@@

```{r}
all_null_decadal_models %>% 
  filter(anova_pval < 0.05) %>% 
  filter(code == "DEUTE") %>% 
  arrange(desc(anova_pval)) %>% 
  select(code, x, sex, model_alt) %>% 
  mutate(
    model_results = map(model_alt, broom::tidy)
  ) %>% 
  select(-model_alt) %>% 
  unnest(model_results) %>% 
  mutate(
    term  = case_when(
      term == "(Intercept)" ~ "Baseline (1980s)",
      term == "factor(decade)1990" ~ "1990s",
      term == "factor(decade)2000" ~ "2000s",
      term == "factor(decade)2010" ~ "2010s",
      TRUE ~ NA_character_
    )
  ) %>% 
  mutate(
    label = glue::glue("{round(estimate, 3)}{case_when(p.value < 0.01 ~ '***', p.value < 0.05 ~ '**', p.value < 0.10 ~ '*', TRUE ~ '')} ({round(std.error, 3)})")
  ) %>% 
  ungroup() %>% 
  select(
    x, Term = term, label
  ) %>% 
  pivot_wider(values_from = label, names_from = x) %>% 
  kable(format = "html", table.attr = "style=\"color: black;\"",caption = "Coefficients of decadal differences, East Germany, females, by starting age (0 = life expectancy at birth; 65 = life expectancy at age 65 years") %>% 
  kableExtra::kable_styling() %>% 
  footnote(general = "The baseline is the average annual rate of improvement in the 1980s. The other terms indicate increments or decrements from this baseline. Statistical significance of coefficients: * if p < 0.10, ** if p < 0.05, *** if p < 0.01"
  )
```

@@

And for Netherlands

```{r}
all_null_decadal_models %>% 
  filter(anova_pval < 0.05) %>% 
  filter(code == "NLD") %>% 
  arrange(desc(anova_pval)) %>% 
  select(code, x, sex, model_alt) %>% 
  mutate(
    model_results = map(model_alt, broom::tidy)
  ) %>% 
  select(-model_alt) %>% 
  unnest(model_results) %>% 
  mutate(
    term  = case_when(
      term == "(Intercept)" ~ "Baseline (1980s)",
      term == "factor(decade)1990" ~ "1990s",
      term == "factor(decade)2000" ~ "2000s",
      term == "factor(decade)2010" ~ "2010s",
      TRUE ~ NA_character_
    )
  ) %>% 
  mutate(
    label = glue::glue("{round(estimate, 3)}{case_when(p.value < 0.01 ~ '***', p.value < 0.05 ~ '**', p.value < 0.10 ~ '*', TRUE ~ '')} ({round(std.error, 3)})")
  ) %>% 
  ungroup() %>% 
  select(
    x, Term = term, label
  ) %>% 
  pivot_wider(values_from = label, names_from = x) %>% 
  kable(format = "html", table.attr = "style=\"color: black;\"",caption = "Coefficients of decadal differences, Netherlands, males, by starting age (0 = life expectancy at birth; 65 = life expectancy at age 65 years") %>% 
  kableExtra::kable_styling() %>% 
  footnote(general = "The baseline is the average annual rate of improvement in the 1980s. The other terms indicate increments or decrements from this baseline. Statistical significance of coefficients: * if p < 0.10, ** if p < 0.05, *** if p < 0.01"
  )
```

So, for the Netherlands, males, e65, the 2000s were a period of exceptionally rapid improvement.
